Thesis - Upwelling and MHW/Cold spells analyses

### 2019-09-08 Write_up/write_up.Rmd
updated

This research study lies within the borders of South Africa along the west coast in particular. The locations of each of the points used within this study can be seen in Fig.1. The South African Coastal Temperature Network was the primary dataset used to specify the various sites along the west coast of South Africa. This dataset consists of coastal seawater temperatures obtained from 129 sites along the entire South African coastline which is approximately 3000km long and is distinct in that it is bordered by the Benguela and Agulhas Currents (e.g.,Roberts 2005; Hutchings et al. 2009). These currents are known to largely influence the seawater temperature and marine life along the South African coastline. The SACTN has collected SST data along this unique coastline from as early as 1972 till present date, with contributions made from several organisations and govermental departments. As a result, temperature within this datasets were obtained by using different methods and instruments, such as hand held thermometers or Underwater temperature reorders( UTRs). The precision of the instruments used vary accordingly, with hand-held thermometers recording temperature at a precision of 0.5°C as opposed to the underwater temperature recorders which have an instrument precision of 0.001°C. The duration and extent of recordings per sites were uneven with the longest time series in the dataset being that of Gordons Bay which started on 13 September 1972 and concluded on 26 January 2017.  During the 1970s, a total of 11 time series began recording. A further 53 entries were added during the 1980s, 34 entries were added during the 1990s, and 18 entries were added during the 2000s. Recordings are still ongoing at many of these sites.

@@

Inorder to create a time series of each of the remotely sensed SST data representing and also comparing the broad-scale temperatures against the *in situ* time series, shore-normal transects were extended from the *in situ* sites a long the west coast of South Africa. Temperature values from the various satellite datasets were extracted at multiple distances from the coast. These distances are 15 x 15 km, 25 X 25 km and 50 X 50 km pixels along the transect (Figure 1). The daily sea surface temperatures obtained from the various remotely sensed datasets within each of the transects were initiated inorder to produce offshore, broad-scale time series that could accurately be used as a dataset to be compared against the *in situ* time series.

@@

At a broad scale, the South African coastline exhibits a large variation in seawater temperature and is divided into four bioregions, each with contrasting temperatures [@smit, @mead]. The western region of the coastline is dominated by the Benguela Current forming an Eastern Boundry Upwelling System (EBUS) [@hutchings]. Seasonal upwelling is controlled by south-easterly trade winds and as a result reflect a distinct temperature variations representing much lower temperatures within the upwelling cells over a fairly narrow continental shelf found from the Cape Peninsula to Cape Columbine. In order to examine upwelling patterns between sites, the appropriate sites along the west coast of South Africa (i.e. sites influenced by the Benguela current) were selected. Since temperature data were not evenly recorded for each of the sites, the entire dataset was narrowed to only those sites that could be adequately compared. Once the sites were minimised, one years worth of data was selected from the SACTN data between the years 1992 – 2015.  This yield a total of six sites along the west coast of South Africa. Hereafter, each of the remotely sensed SST datasets were matched to the SACTN dataset. This was done for two reasons, to find the same points at which temperature was collected for the SACTN sites and secondly, to confirm that the dates and time that temperature were collected correspond accordingly.

@@

knitr::include_graphics("../Figures/map_fixed.png")
```

>>>>

This research study lies within the Benguela region along the west coast of South Africa. Our study sites covered the full range of this coastal region (Fig.1). The South African Coastal Temperature Network (SACTN) data serves as the *in situ* temperature dataset used in this study. This dataset consists of coastal seawater temperatures obtained from 129 sites along the South African coastline which is approximately 3000km long and is distinct in that it is bordered by the Benguela and Agulhas Currents (e.g.,Roberts 2005; Hutchings et al. 2009). These currents are known to largely influence the seawater temperature and marine life along the South African coastline. The SACTN has collected SST data along this unique coastline from as early as 1972 till present date, with contributions made from several organisations and govermental departments. As a result, temperature within this datasets were obtained by using different methods and instruments, such as hand held thermometers or Underwater temperature reorders( UTRs). The precision of the instruments used vary accordingly, with hand-held thermometers recording temperature at a precision of 0.5°C as opposed to the underwater temperature recorders which have an instrument precision of 0.001°C. The duration and extent of recordings per sites were uneven with the longest time series in the dataset being that of Gordons Bay which started on 13 September 1972 and concluded on 26 January 2017.  During the 1970s, a total of 11 time series began recording. A further 53 entries were added during the 1980s, 34 entries were added during the 1990s, and 18 entries were added during the 2000s. Recordings are still ongoing at many of these sites.

@@

Inorder to create a time series of each of the remotely sensed SST data and also to compare whether the same upwelling patterns exist at the broad-scale against the *in situ* time series, shore-normal transects were extended from the *in situ* sites a long the west coast of South Africa. Temperature values from the various satellite datasets were extracted at multiple distances from the coast. These distances are 15 x 15 km, 25 X 25 km and 50 X 50 km pixels along the transect (Fig. 1). The daily sea surface temperatures obtained from the various remotely sensed datasets within each of the transects were initiated inorder to produce offshore, broad-scale time series that could accurately be used as a dataset to be compared against the *in situ* time series.

@@

At a broad scale, the South African coastline exhibits a large variation in seawater temperature and is divided into four bioregions, each with contrasting temperatures [@smit, @mead]. The western region of the coastline is dominated by the Benguela Current forming an Eastern Boundry Upwelling System (EBUS) [@hutchings] which provides a natural labratory for this study. Seasonal upwelling is controlled by south-easterly trade winds and as a result reflect a distinct temperature variations representing much lower temperatures within the upwelling cells over a fairly narrow continental shelf found from the Cape Peninsula to Cape Columbine. This region represents monthly mean temperatures at generally intermediate at cold and warm temperate [@luning]. In order to examine upwelling patterns between sites, the appropriate sites along the west coast of South Africa (i.e. sites influenced by the Benguela current) were selected. Since temperature data were not evenly recorded for each of the sites, the entire dataset was narrowed to only those sites with a time series of longer than 30 years. This yield a total of five sites within the Benguela region.

@@

knitr::include_graphics("../Figures/site_map.png")
```

# Defining and calculating the occurance of upwelling

Inorder to determine whether or not upwelling signals are present at different distances from the coastline and to test which dataset shows clear, visible representation of these events, it was necessary to define when upwelling is occuring. For this research study we aimed to determine a threshold value for which we could identify that upwelling was occuring at a particular time. Given that upwelling is primarily caused by alongshore, equator ward winds, it was necessary to consider temperature and wind values when determining a threshold for temperature to be experiencing an upwelling event. We then decided to make use of the following equation $$ UpwellingIndex = μ{(Cosθ − coastal angle)} $$

In this equation μ represents the wind speed (m/s) and θ represents the wind direction which is measured in degrees. The coastal angle is the angle of each of the sites perpendicular to the coastline. Thereafter, the 'exceedance' function within the heatwaveR package was used to determine the consecutive number of days at or above which the upwelling index value is ment to be. Here it was specified that the minimum duration for upwelling is 1 day with a threshold value obtained from the upwelling indeces as 1. Inorder to obtain a more approved threshold value, a percentile value was determined. To determine changes within the upwelling values, the 'detect event' fuction within the heatwaveR package was used. Thereafter, to determine the upwelling event metrics ( frequency and intensity) the detect event function was used as a first filter with a threshold of 25th percentile. Values obtained from the upwelling indeces were used as a second threshold. An R script (https://github.com/AmierohAbrahams/Upwelling_MCS) was used to further used to calculate the individual upwelling events for both the *in situ* and remotely sensed SST data. Because upwelling events were calculated relative to percentile exceedences, rather than an absolute definition such as periods with temperatures above a fixed temperature threshold, upwelling events occured any time of the year with upwelling shown to be more dominant during spring and summer months.
